\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-10-05 papers/methods/paper.tex
Initial Tam's comments

Two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SNIa) has increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, chambers2016panstarrs, sako2018sdss}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014, Scolnic2017} have been combined with complementary probes of large scale structure and the CMB. For a recent review, see \citet{Huterer2018}. {\green While these efforts have reduced the uncertainty on the equation-of-state of dark energy by more than a factor of two, it is still consistent with a cosmological constant and the nature of dark energy remains an unsolved mystery.}

In attempts to tease out the nature of dark energy, active and planned surveys are {\green continually growing in size and scale}. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has observed thousands of Type Ia supernova, attaining both spectroscopic and photometric identified samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will observe scores of thousands of photometrically classified supernovae. {\green Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as we will required reduced systematic uncertainties to fully utilised increase statistics} \citep{Betoule2014, Scolnic2017}.

@@

{\green Forefront among these challenges is our ignorance of the underlying Type Ia population -- its intrinsic dispersion}. Ideally, analysis of the underlying population would make use of an ensemble of time-series spectroscopy to characterise the diversity of Type Ia supernovae, however this data is difficult to obtain, and recent efforts to quantify the population draw inference from photometric measurements. We utilise two dispersion models in this work. The {\green first is based on the} \citet[][hereafter denoted {\gten}]{Guy2010} scatter model, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second, {\green based on the} {\celeven} model, is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation. {\green These two broadband scatter models are converted to spectral energy distribution models for use in simulations in \citet{Kessler13}.} As the SALT2 model does not include full treatment of intrinsic dispersion, each scatter model results in different biases in $m_B$, $x_1$ and $c$ when fitting the SALT2 model to light curve observations, and results in increased uncertainty on the summary statistics that is not encapsulated in the reported covariance $\mathcal{C}$. These two scatter models are currently {\green assumed} to span the possible range of scatter in the underlying supernova population. We have insufficient information to prefer one model over the other, and thus we have to account for both possible scatter models.

The underlying supernova population is further complicated by the presence of outliers. Non-Ia supernovae often trigger transient follow-up in surveys and can easily be mistaken for Type Ia supernovae {\green and represent outliers from the standardised SN Ia sample.} This contamination is not just a result of non-SNIa being observed, but can also arise from host galaxy misidentification causing incorrect redshifts. Depending on the cuts {\green optimising the} ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification \citep{Gupta2016} and results in a broad population of outliers. For spectroscopic surveys, where both supernova type and redshift can be confirmed through the supernova spectra, this outlier population is negligible. However, for photometric surveys, which do not have the spectroscopic confirmation, it is one of the largest challenges; how to model, fit and correct for contaminants.

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SNIa, datasets are commonly formed from the discoveries of multiple surveys. However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

\subsubsection{Observables}
Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a flux measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty on those values encoded in the covariance matrix $\cov$. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy {\green stellar }mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of each supernovae being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

For future use with photometrically classified surveys, we include a simplistic outlier population model. We follow {\rubin} \citep[and therefore ][]{Kunz2007} by implementing a Gaussian mixture model. For surveys with low rates of contamination, it is not possible to fit a contamination population, and the mean of the outlier population has been fixed to the SNIa population in prior works. However, with with the increased number of contaminants expected in the DES photometric sample, we seek a more physically motivated outlier population.  We find that an acceptable parametrisation is to model the outlier population with a mean absolute magnitude of $\langle M_B^{\rm outl} \rangle = \langle M_B \rangle + \delta_{M_B}^{\rm outl}$, where $\delta_{M_B}^{\rm outl}$ is constrained to be positive, or even to be greater than a small positive number to reduce degeneracy between the two populations. We set the population width to $\sigma^{\rm outl} = 1$ in $M_B$, $x_1$ and $c$. The probability of each supernova falling into either population is determined by the observed Type Ia probability $\hat{p}$ discussed in the previous section. For the spectroscopic survey, we set this to unity. For the photometric proof-of-concept we provide an accurate probability estimate. Further investigation on the effect of inaccurate {\green estimates will be left for} future improvements during the analysis of the DES photometric sample.

@@

It is now well known that host galaxy environment is significantly correlated with supernova properties. The latest sample of over 1300 spectroscopically confirmed Type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014} invoke a step function such that $\Delta M = 0.08 \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function, $M$ is the galaxy mass in solar masses and $0.08$ represents the size of the magnitude step. The scale of this step function varies from analysis to analysis, with the 0.08 value shown previously sourced from \cite{Sullivan2010} and used in \citet{Betoule2014}. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:

@@

With the well sampled approximation we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), the correction \textit{per SNIa} becomes

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{efficiency.pdf}
	\end{center}
	\caption{The efficiency for supernova discovery at an arbitrary redshift. Shown in both panels in dashed blue is the SNIa population distribution, which takes the form of a normal distribution. The top panel shows a CDF based survey efficiency (green dotted line), whilst the bottom panel shows a skew normal based survey efficiency (red dotted line), as functions of apparent magnitude. The survey efficiency, given the SNIa population, is shown as a solid line in both panels, and the probability of observing a SNIa is found by integrating over the population detection efficiency as described in equation \eqref{eq:w1}, and has been shown by shading the area integrated. This area is what is analytically given by equations \eqref{eq:seldes} and \eqref{eq:sellow-$z$}.}
	\label{fig:efficiency}
\end{figure}

@@

\begin{enumerate}[1.]
	\item Draw a redshift from a power law distribution. For the low redshift survey this is $\mathcal{U}(0.0004, 0.01)^{0.5}$, and for the DES-like survey this is $\mathcal{U}(0.008, 1.0)^{0.3}$.
	\item Draw a random mass probability from $\mathcal{U}(0, 1)$ and calculate the mass-brightness correction using $\delta(0) = 0.08$, $\delta(0)/\delta(\infty) = 0.5$, and equation \eqref{eq:mass}.
	\item Draw an absolute magnitude, stretch and colour from the respective distributions $\mathcal{N}(-19.3, 0.1)$, $\mathcal{N}(0, 1)$, $\mathcal{N}(0, 0.1)$. {\green Calculate the true absolute magnitude using Equation \eqref{eq:l2}.}
	\item Calculate $\mu(z)$ given the drawn redshift and cosmological parameters $\Omega_m = 0.3$, $w = -1$ under Flat $\Lambda$CDM cosmology. Use this to determine the true apparent magnitude of the object $m_B$.
	\item Determine if the SNIa is detected using detection probability $P(S|m_B) = \mathcal{N}^{\rm skew}(13.72, 1.35, 5.87)$ for the low redshift survey (numeric values obtained by fitting to existing low redshift data). For the DES-like survey, accept with probability $P(S|m_B) = \Phi^C(23.14, 0.5)$. Repeat from step one until we have a supernova that passes. {\green We use realistic values for the selection probability to ensure our model is numerically stable with highly skewed selection functions.}
	\item Add independent, Gaussian observational error onto the true $m_B$, $x_1$, $c$ using Gaussian widths of $0.04$, $0.2$, $0.03$ respectively (following the mean uncertainty for DES-like SNANA simulations). Add extra colour uncertainty in quadrature of $\kappa_0 + \kappa_1 z$, where $\kappa_0 = \kappa_1 = 0.03$.
\end{enumerate}

@@

The importance of validating models using high-precision statistics gained by performing fits to hundreds of data realisations cannot be overstated, however this validation is lacking in many earlier BHM models for supernova cosmology. We have validated this model against many realisations of simplistic simulations with well-known and well-defined statistics, and found no cosmological bias. When validating using SNANA simulations, we find evidence of cosmological bias which is traced back to light curve fits reporting biased observables and incorrect covariance. Allowing fully parametrised corrections on observed supernovae summary statistics introduces too many degrees of freedom and is found to make cosmology fits too weak. Allowing simulation based corrections to vary in strength is found to give minor reductions in $w$ bias, however the uncertainty on the intrinsic scatter model itself limits the efficacy of the bias corrections. For the data size represented in the DES three-year spectroscopic survey, the determined biases should be sub-dominant to other sources of uncertainty, however this cannot be expected for future analyses with larger datasets. Stricter bias corrections calculated from simulations are required to reduce bias. Ideally, this would include further work on the calculation of intrinsic dispersion of the Type Ia supernovae population such that we can better characterise this bias.

>>>>

Two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed type Ia supernovae (SN~Ia) has increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, chambers2016panstarrs, sako2018sdss}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014, Scolnic2017} have been combined with complementary probes of large scale structure and the CMB. For a recent review, see \citet{Huterer2018}. {\green While these efforts have reduced the uncertainty on the equation-of-state of dark energy by more than a factor of two, it is still consistent with a cosmological constant and the nature of dark energy remains an unsolved mystery.}

In attempts to tease out the nature of dark energy, active and planned surveys are {\green continually growing in size and scale}. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has observed thousands of type Ia supernova, attaining both spectroscopic and photometric identified samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will observe scores of thousands of photometrically classified supernovae. {\green Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as we will required reduced systematic uncertainties to fully utilised increase statistics} \citep{Betoule2014, Scolnic2017}.

@@

{\green Forefront among these challenges is our ignorance of the underlying type Ia population -- its intrinsic dispersion}. Ideally, analysis of the underlying population would make use of an ensemble of time-series spectroscopy to characterise the diversity of type Ia supernovae, however this data is difficult to obtain, and recent efforts to quantify the population draw inference from photometric measurements. We utilise two dispersion models in this work. The {\green first is based on the} \citet[][hereafter denoted {\gten}]{Guy2010} scatter model, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second, {\green based on the} {\celeven} model, is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation. {\green These two broadband scatter models are converted to spectral energy distribution models for use in simulations in \citet{Kessler13}.} As the SALT2 model does not include full treatment of intrinsic dispersion, each scatter model results in different biases in $m_B$, $x_1$ and $c$ when fitting the SALT2 model to light curve observations, and results in increased uncertainty on the summary statistics that is not encapsulated in the reported covariance $\mathcal{C}$. These two scatter models are currently {\green assumed} to span the possible range of scatter in the underlying supernova population. We have insufficient information to prefer one model over the other, and thus we have to account for both possible scatter models.

The underlying supernova population is further complicated by the presence of outliers. Non-Ia supernovae often trigger transient follow-up in surveys and can easily be mistaken for type Ia supernovae {\green and represent outliers from the standardised SN Ia sample.} This contamination is not just a result of non-SN~Ia being observed, but can also arise from host galaxy misidentification causing incorrect redshifts. Depending on the cuts {\green optimising the} ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification \citep{Gupta2016} and results in a broad population of outliers. For spectroscopic surveys, where both supernova type and redshift can be confirmed through the supernova spectra, this outlier population is negligible. However, for photometric surveys, which do not have the spectroscopic confirmation, it is one of the largest challenges; how to model, fit and correct for contaminants.

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SN~Ia, datasets are commonly formed from the discoveries of multiple surveys. However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

\subsubsection{Observables}
Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a flux measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty on those values encoded in the covariance matrix $\cov$. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy {\green stellar }mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of each supernovae being a type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

For future use with photometrically classified surveys, we include a simplistic outlier population model. We follow {\rubin} \citep[and therefore ][]{Kunz2007} by implementing a Gaussian mixture model. For surveys with low rates of contamination, it is not possible to fit a contamination population, and the mean of the outlier population has been fixed to the SN~Ia population in prior works. However, with with the increased number of contaminants expected in the DES photometric sample, we seek a more physically motivated outlier population.  We find that an acceptable parametrisation is to model the outlier population with a mean absolute magnitude of $\langle M_B^{\rm outl} \rangle = \langle M_B \rangle + \delta_{M_B}^{\rm outl}$, where $\delta_{M_B}^{\rm outl}$ is constrained to be positive, or even to be greater than a small positive number to reduce degeneracy between the two populations. We set the population width to $\sigma^{\rm outl} = 1$ in $M_B$, $x_1$ and $c$. The probability of each supernova falling into either population is determined by the observed type Ia probability $\hat{p}$ discussed in the previous section. For the spectroscopic survey, we set this to unity. For the photometric proof-of-concept we provide an accurate probability estimate. Further investigation on the effect of inaccurate {\green estimates will be left for} future improvements during the analysis of the DES photometric sample.

@@

It is now well known that host galaxy environment is significantly correlated with supernova properties. The latest sample of over 1300 spectroscopically confirmed type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014} invoke a step function such that $\Delta M = 0.08 \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function, $M$ is the galaxy mass in solar masses and $0.08$ represents the size of the magnitude step. The scale of this step function varies from analysis to analysis, with the 0.08 value shown previously sourced from \cite{Sullivan2010} and used in \citet{Betoule2014}. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:

@@

With the well sampled approximation we can remove the redshift integral in Eq \eqref{eq:w1} and replace it with a correction for each observed supernova. For the error function (denoted with the subscript `CDF') and skew normal selection functions respectively (denoted with a subscript `Skew'), the correction \textit{per SN~Ia} becomes

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{efficiency.pdf}
	\end{center}
	\caption{The efficiency for supernova discovery at an arbitrary redshift. Shown in both panels in dashed blue is the SN~Ia population distribution, which takes the form of a normal distribution. The top panel shows a CDF based survey efficiency (green dotted line), whilst the bottom panel shows a skew normal based survey efficiency (red dotted line), as functions of apparent magnitude. The survey efficiency, given the SN~Ia population, is shown as a solid line in both panels, and the probability of observing a SN~Ia is found by integrating over the population detection efficiency as described in equation \eqref{eq:w1}, and has been shown by shading the area integrated. This area is what is analytically given by equations \eqref{eq:seldes} and \eqref{eq:sellow-$z$}.}
	\label{fig:efficiency}
\end{figure}

@@

\begin{enumerate}[1.]
	\item Draw a redshift from a power law distribution. For the low redshift survey this is $\mathcal{U}(0.0004, 0.01)^{0.5}$, and for the DES-like survey this is $\mathcal{U}(0.008, 1.0)^{0.3}$.
	\item Draw a random mass probability from $\mathcal{U}(0, 1)$ and calculate the mass-brightness correction using $\delta(0) = 0.08$, $\delta(0)/\delta(\infty) = 0.5$, and equation \eqref{eq:mass}.
	\item Draw an absolute magnitude, stretch and colour from the respective distributions $\mathcal{N}(-19.3, 0.1)$, $\mathcal{N}(0, 1)$, $\mathcal{N}(0, 0.1)$. {\green Calculate the true absolute magnitude using Equation \eqref{eq:l2}.}
	\item Calculate $\mu(z)$ given the drawn redshift and cosmological parameters $\Omega_m = 0.3$, $w = -1$ under Flat $\Lambda$CDM cosmology. Use this to determine the true apparent magnitude of the object $m_B$.
	\item Determine if the SN~Ia is detected using detection probability $P(S|m_B) = \mathcal{N}^{\rm skew}(13.72, 1.35, 5.87)$ for the low redshift survey (numeric values obtained by fitting to existing low redshift data). For the DES-like survey, accept with probability $P(S|m_B) = \Phi^C(23.14, 0.5)$. Repeat from step one until we have a supernova that passes. {\green We use realistic values for the selection probability to ensure our model is numerically stable with highly skewed selection functions.}
	\item Add independent, Gaussian observational error onto the true $m_B$, $x_1$, $c$ using Gaussian widths of $0.04$, $0.2$, $0.03$ respectively (following the mean uncertainty for DES-like SNANA simulations). Add extra colour uncertainty in quadrature of $\kappa_0 + \kappa_1 z$, where $\kappa_0 = \kappa_1 = 0.03$.
\end{enumerate}

@@

The importance of validating models using high-precision statistics gained by performing fits to hundreds of data realisations cannot be overstated, however this validation is lacking in many earlier BHM models for supernova cosmology. We have validated this model against many realisations of simplistic simulations with well-known and well-defined statistics, and found no cosmological bias. When validating using SNANA simulations, we find evidence of cosmological bias which is traced back to light curve fits reporting biased observables and incorrect covariance. Allowing fully parametrised corrections on observed supernovae summary statistics introduces too many degrees of freedom and is found to make cosmology fits too weak. Allowing simulation based corrections to vary in strength is found to give minor reductions in $w$ bias, however the uncertainty on the intrinsic scatter model itself limits the efficacy of the bias corrections. For the data size represented in the DES three-year spectroscopic survey, the determined biases should be sub-dominant to other sources of uncertainty, however this cannot be expected for future analyses with larger datasets. Stricter bias corrections calculated from simulations are required to reduce bias. Ideally, this would include further work on the calculation of intrinsic dispersion of the type Ia supernovae population such that we can better characterise this bias.
